---
title: "In-class Exercise 7"
author: "Kenneth Yong"
---

# Getting Started

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

## Importing the Data

```{r}
ts_data <- read_csv(
  "data/visitor_arrivals_by_air.csv")
```

In the code chunk below, dmy() of lubridate package is used to convert data type of Month-Year field from Character to Date.

```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

## Conventional base ts object versus tibble object

tibble object

```{r}
ts_data
```

ts object

```{r}
ts_data_ts <- ts(ts_data)       
head(ts_data_ts)
```

## Converting tibble object to tsibble object

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
```

```{r}
ts_tsibble
```

# Visualising Time-series Data

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## Visualising single time-series: ggplot2 methods

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

## Plotting multiple time-series data with ggplot2 methods

```{r}
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

```{r}
#| echo: false
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,
             scales = "free_y") +
  theme_bw()
```

# Visual Analysis of Time-series Data

```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## Visual Analysis of Seasonality with Seasonal Plot

```{r}
tsibble_longer %>%
  filter(Country == "Italy" |
         Country == "Vietnam" |
         Country == "United Kingdom" |
         Country == "Germany") %>% 
  gg_season(Arrivals)
```

## Visual Analysis of Seasonality with Cycle Plot

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  autoplot(Arrivals) + 
  facet_grid(Country ~ ., scales = "free_y")
```

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```

# Time series decomposition

## Single time series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  ACF(Arrivals) %>% 
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  PACF(Arrivals) %>% 
  autoplot()
```

## Multiple time-series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  ACF(Arrivals) %>%
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  PACF(Arrivals) %>%
  autoplot()
```

## Composite plot of time series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  gg_tsdisplay(Arrivals)
```

# Visual STL Diagnostics

## Visual STL diagnostics with feasts

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

## Classical Decomposition with feasts

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(
    classical_decomposition(
      Arrivals, type = "additive")) %>%
  components() %>%
  autoplot()
```

# Visual Forecasting

## Time Series Data Sampling

```{r}
vietnam_ts <- tsibble_longer %>%
  filter(Country == "Vietnam") %>% 
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training"))
```

```{r}
vietnam_train <- vietnam_ts %>%
  filter(`Month-Year` < "2019-01-01")
```
