---
title: "Take-home Exercise 3"
author: "Kenneth Yong"
---

## Overview

We use `geofacet` to visualize weather data for Singapore across multiple stations in a geographically meaningful layout.

## Data Preparation

The data preparation process is documented [here](https://weatherxplore.netlify.app/data_prep).

## Load Required Packages

```{r}
pacman::p_load(tidyverse, geofacet, ggplot2, readr)
```

## Import the Dataset

```{r}
weather_data <-read_csv("data/weather_data_clean.csv")
```

## Prepare Custom Geofacet Grid

We created a custom grid layout to approximate the geographical positioning of 19 weather stations in Singapore.

```{r}
custom_grid <- data.frame(
  name = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi", "Tengah",
           "Jurong Island", "Tuas South", "Sentosa Island", "Marina Barrage",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  code = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi", "Tengah",
           "Jurong Island", "Tuas South", "Sentosa Island", "Marina Barrage",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  row = c(1, 1, 1, 1,
          2, 2, 2, 2,
          3, 3, 3, 3,
          4, 4, 4, 4,
          5, 5, 5),
  col = c(2, 1, 3, 4,
          2, 1, 3, 4,
          1, 2, 3, 4,
          1, 2, 3, 4,
          1, 2, 3),
  stringsAsFactors = FALSE
)
```

## Visualize the Data with `facet_geo()`

### Monthly Mean Temperature:

```{r}
# Aggregate to monthly means
weather_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station",
       x = "Month", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Yearly Mean Temperature:

```{r}
# Aggregate to yearly means
weather_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_yearly, aes(x = Year, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Mean Temperature by Station",
       x = "Year", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```