---
title: "Take-home Exercise 3"
author: "Kenneth Yong"
---

## Overview

We use `geofacet` to visualize weather data for Singapore across multiple stations in a geographically meaningful layout.

## Data Preparation

The data preparation process is documented [here](https://weatherxplore.netlify.app/data_prep).

## Load Required Packages

```{r}
pacman::p_load(tidyverse, geofacet, ggplot2, readr)
```

## Import the Dataset

```{r}
weather_data <-read_csv("data/weather_data_updated.csv")
```

## Prepare Custom Geofacet Grid

We created a custom grid layout to approximate the geographical positioning of 19 weather stations in Singapore.

```{r}
custom_grid <- data.frame(
  name = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  code = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  row = c(2, 1, 2, 2,
          3, 3, 4, 3,
          2, 3, 4, 5,
          4, 5, 4, 4,
          3),
  col = c(3, 3, 4, 5,
          4, 3, 4, 5,
          2, 2, 2, 2,
          1, 3, 3, 5,
          6),
  stringsAsFactors = FALSE
)
```

## Visualize the Data with `facet_geo()`

### Monthly Mean Temperature:

```{r}
# Aggregate to monthly means
weather_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station",
       x = "Month", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Monthly Rainfall Total:

```{r}
# Aggregate to monthly means
weather_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Daily Rainfall Total (mm)` = mean(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_monthly, aes(x = Month, y = `Daily Rainfall Total (mm)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Rainfall Total by Station",
       x = "Month", y = "Monthly Rainfall Total (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Yearly Mean Temperature:

```{r}
# Aggregate to yearly means
weather_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_yearly, aes(x = Year, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Mean Temperature by Station",
       x = "Year", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Filter for 2024 and aggregate to monthly means
weather_2024_monthly <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

# Geofacet plot for 2024
ggplot(weather_2024_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station (2024)",
       x = "Month", y = "Mean Temperature (°C)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```