---
title: "Take-home Exercise 3"
author: "Kenneth Yong"
---

## Overview
We use `geofacet` to visualize weather data for Singapore across multiple stations in a geographically meaningful layout.

## Step 1: Load Required Packages

```{r}
pacman::p_load(tidyverse, geofacet, ggplot2, readr)
```

## Step 2: Import the Dataset

```{r}
weather_data <-read_csv("data/weather_data_clean.csv")
```

## Step 3: Prepare Custom Geofacet Grid
We created a custom grid layout to approximate the geographical positioning of 19 weather stations in Singapore.

```{r}
custom_grid <- data.frame(
  name = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi", "Tengah",
           "Jurong Island", "Tuas South", "Sentosa Island", "Marina Barrage",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  code = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi", "Tengah",
           "Jurong Island", "Tuas South", "Sentosa Island", "Marina Barrage",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  row = c(1, 1, 1, 1,
          2, 2, 2, 2,
          3, 3, 3, 3,
          4, 4, 4, 4,
          5, 5, 5),
  col = c(2, 1, 3, 4,
          2, 1, 3, 4,
          1, 2, 3, 4,
          1, 2, 3, 4,
          1, 2, 3),
  stringsAsFactors = FALSE
)
```

## Step 4: Visualize the Data with `facet_geo()`
Assuming your dataset is named `weather_data` and includes a `Station`, `Date`, and `Mean Temperature (°C)` column:

```{r}
ggplot(weather_data, aes(x = Date, y = `Mean Temperature (°C)`)) +
  geom_line(color = "#2c3e50") +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Mean Temperature Over Time by Station",
       x = "Date", y = "Mean Temperature (°C)")
```