---
title: "Take-home Exercise 3"
author: "Kenneth Yong"
---

# Overview

We use `geofacet` to visualize weather data for Singapore across multiple stations in a geographically meaningful layout.

# Data Preparation

The data preparation process is documented [here](https://weatherxplore.netlify.app/data_prep).

## Load Required Packages

```{r}
pacman::p_load(tidyverse, geofacet, ggplot2, readr, dplyr, sf, sp, gstat, stars, raster, tmap, lubridate)
```

## Import the Dataset

```{r}
weather_data <-read_csv("data/weather_data_cleaned.csv")
```

## Prepare Custom Geofacet Grid

We created a custom grid layout to approximate the geographical positioning of 19 weather stations in Singapore.

```{r}
custom_grid <- data.frame(
  name = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  code = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  row = c(2, 1, 2, 2,
          3, 3, 4, 3,
          2, 3, 4, 5,
          4, 5, 4, 4,
          3),
  col = c(3, 3, 4, 5,
          4, 3, 4, 5,
          2, 2, 2, 2,
          1, 3, 3, 5,
          6),
  stringsAsFactors = FALSE
)
```

# Visualize the Data with `facet_geo()`

Geofaceting with line graphs and bar charts.

## 1. Monthly Mean Temperature:

```{r}
# Aggregate to monthly means
weather_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station",
       x = "Month", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Monthly Total Rainfall:

```{r}
# Aggregate to monthly total rainfall
rainfall_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Total Monthly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet plot
ggplot(rainfall_monthly, aes(x = Month, y = `Total Monthly Rainfall (mm)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Total Monthly Rainfall by Station",
       x = "Month", y = "Total Monthly Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3. Mean Temperature Trend (2020-2024):

```{r}
# Aggregate to yearly means
weather_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_yearly, aes(x = Year, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Mean Temperature by Station",
       x = "Year", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Total Rainfall Trend (2020-2024):

```{r}
# Aggregate to yearly total rainfall
rainfall_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Total Yearly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart of yearly total rainfall
ggplot(rainfall_yearly, aes(x = factor(Year), y = `Total Yearly Rainfall (mm)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Total Rainfall by Station",
       x = "Year", y = "Total Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Mean Monthly Temperature by Station (2024)

Filter for desired year (eg. 2024).

```{r}
# Filter for 2024 and aggregate to monthly means
weather_2024_monthly <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

# Geofacet plot for 2024
ggplot(weather_2024_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station (2024)",
       x = "Month", y = "Mean Temperature (°C)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 6. Monthly Total Rainfall by Station (2024)

Filter for desired year (eg. 2024).

```{r}
# Aggregate to monthly total rainfall for 2024
rainfall_2024 <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Daily Rainfall Total (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart with abbreviated month labels
ggplot(rainfall_2024, aes(x = Month, y = `Daily Rainfall Total (mm)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Total Rainfall by Station (2024)",
       x = "Month", y = "Total Rainfall (mm)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Aggregate to monthly total rainfall for 2024
rainfall_2024 <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Daily Rainfall Total (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart with abbreviated month labels
ggplot(rainfall_2024, aes(x = Month, y = `Daily Rainfall Total (mm)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Total Rainfall by Station (2024)",
       x = "Month", y = "Total Rainfall (mm)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 7. Monthly Climatology (Average Across Years)

This plot answers the question: "What’s a typical temperature like in each month for each part of Singapore?"

```{r}
weather_climatology <- weather_data %>%
  filter(year(Date) %in% 2020:2024) %>%
  mutate(Month_Label = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Station, Region, Month_Label) %>%
  summarise(`Avg Temp (2020–2024)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_climatology, aes(x = Month_Label, y = `Avg Temp (2020–2024)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Climatology (2020–2024) by Station",
       x = "Month", y = "Avg Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))
```

## 8. Extreme Rainfall Events Count

Filter for 2024 (or a desired range), then define an extreme rainfall threshold (e.g., \> 80mm). This plot shows the frequency of days with heavy rainfall (\>80mm, for example).

```{r}

# Define threshold
extreme_threshold <- 80

# Prepare base data with month labels
extreme_rainfall <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month"),
         Month_Label = factor(format(Month, "%b"), levels = month.abb)) %>%
  filter(`Daily Rainfall Total (mm)` > extreme_threshold) %>%
  group_by(Station, Region, Month_Label) %>%
  summarise(`Extreme Rainfall Days` = n(), .groups = "drop") %>%
  complete(Station, Region, Month_Label = factor(month.abb, levels = month.abb), fill = list(`Extreme Rainfall Days` = 0))

# Plot
ggplot(extreme_rainfall, aes(x = Month_Label, y = `Extreme Rainfall Days`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Extreme Rainfall Events by Station (2024)",
       subtitle = paste0("Days with rainfall > ", extreme_threshold, "mm"),
       x = "Month", y = "Number of Extreme Rainfall Days") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))

```

## 9. Dry vs Wet Days Per Month

```{r}
# Categorize days as 'Dry' or 'Wet' and count per month
dry_wet_days <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month"),
         Month_Label = factor(format(Month, "%b"), levels = month.abb),
         RainType = ifelse(`Daily Rainfall Total (mm)` > 0, "Wet", "Dry")) %>%
  group_by(Station, Region, Month_Label, RainType) %>%
  summarise(Days = n(), .groups = "drop")

# Geofacet stacked bar chart
ggplot(dry_wet_days, aes(x = Month_Label, y = Days, fill = RainType)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_manual(values = c("Dry" = "skyblue", "Wet" = "steelblue")) +
  theme_minimal() +
  labs(title = "Dry vs Wet Days Per Month by Station (2024)",
       x = "Month", y = "Number of Days", fill = "Day Type") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))

```

## 10. Rainfall Anomalies (Z-score) plot

```{r}
# Step 1: Aggregate to monthly total rainfall
rainfall_monthly <- weather_data %>%
  mutate(Year = year(Date),
         Month = floor_date(Date, "month"),
         Month_Label = month(Month, label = TRUE, abbr = TRUE)) %>%
  group_by(Station, Region, Month, Year, Month_Label) %>%
  summarise(`Monthly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Step 2: Calculate monthly mean & SD for each Station + Month_Label (climatology)
rainfall_stats <- rainfall_monthly %>%
  group_by(Station, Month_Label) %>%
  summarise(
    Mean_Rainfall = mean(`Monthly Rainfall (mm)`, na.rm = TRUE),
    SD_Rainfall = sd(`Monthly Rainfall (mm)`, na.rm = TRUE),
    .groups = "drop"
  )

# Step 3: Join and compute Z-score
rainfall_anomalies <- rainfall_monthly %>%
  left_join(rainfall_stats, by = c("Station", "Month_Label")) %>%
  mutate(Z_Score = (`Monthly Rainfall (mm)` - Mean_Rainfall) / SD_Rainfall)

# Step 4: Plot Z-scores for a specific year (e.g., 2024)
rainfall_anomalies_2024 <- rainfall_anomalies %>%
  filter(Year == 2024)

# Plot
ggplot(rainfall_anomalies_2024, aes(x = Month_Label, y = Z_Score, fill = Z_Score > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"), labels = c("Drier", "Wetter")) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Rainfall Anomalies by Station (2024)",
       subtitle = "Z-score of Monthly Rainfall vs Historical Mean",
       x = "Month", y = "Rainfall Z-score", fill = "Anomaly") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))
```

## 11. Hextile Map (Total Rainfall for 2024)

```{r}
rainfall_total_2024 <- weather_data %>%
  filter(lubridate::year(Date) == 2024) %>%
  group_by(Station, Region) %>%
  summarise(`Total Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

ggplot(rainfall_total_2024, aes(x = 1, y = 1, fill = `Total Rainfall (mm)`)) +
  geom_tile(color = "white") +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_void() +
  labs(title = "Total Rainfall by Station (2024)",
       fill = "Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        legend.position = "bottom")
```

## 12. Hextile Map (Mean Temperature for 2024)

```{r}
temperature_mean_2024 <- weather_data %>%
  filter(lubridate::year(Date) == 2024) %>%
  group_by(Station, Region) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

library(ggplot2)
library(geofacet)

ggplot(temperature_mean_2024, aes(x = 1, y = 1, fill = `Mean Temperature (°C)`)) +
  geom_tile(color = "white") +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_gradient(low = "mistyrose", high = "darkred") +
  theme_void() +
  labs(title = "Mean Temperature by Station (2024)",
       fill = "Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        legend.position = "bottom")
```

# References

<https://ryanhafen.com/blog/geofacet/>
