---
title: "Take-home Exercise 3"
author: "Kenneth Yong"
---

# Overview

To effectively explore and compare weather patterns across different locations in Singapore, this module leverages both geofacet and spatial interpolation techniques.

# Data Preparation

The data preparation process is documented [here](https://weatherxplore.netlify.app/data_prep).

## Additional Preparation for Spatial Interpolation

Download KML file Master Plan 2019 Subzone Boundary (No Sea) (KML) from data.gov.sg and convert to SHP.

``` r
library(sf)
kml_data <- st_read("data/geospatial/MPSZ-2019.kml")
st_write(kml_data, "data/geospatial/MPSZ-2019.shp")
```

## Load Required Packages

```{r}
pacman::p_load(tidyverse, geofacet, ggplot2, readr, sf, terra, gstat, automap, tmap, viridis, dplyr, lubridate)
```

## Import the Dataset

```{r}
weather_data <-read_csv("data/weather_data_cleaned.csv")
```

## Prepare Custom Geofacet Grid

We created a custom grid layout to approximate the geographical positioning of 19 weather stations in Singapore.

```{r}
custom_grid <- data.frame(
  name = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  code = c("Admiralty", "Sembawang", "Seletar", "Pulau Ubin",
           "Ang Mo Kio", "Newton", "Tai Seng", "Paya Lebar",
           "Choa Chu Kang (South)", "Jurong (West)", "Clementi",
           "Jurong Island", "Tuas South", "Sentosa Island",
           "Pasir Panjang", "East Coast Parkway", "Changi"),
  row = c(2, 1, 2, 2,
          3, 3, 4, 3,
          2, 3, 4, 5,
          4, 5, 4, 4,
          3),
  col = c(3, 3, 4, 5,
          4, 3, 4, 5,
          2, 2, 2, 2,
          1, 3, 3, 5,
          6),
  stringsAsFactors = FALSE
)
```

# Geofacet

Geofacet is a visualization technique that arranges multiple small charts in a grid that mirrors the geographic layout of regions, making it easier to compare data across locations while preserving spatial context. In this module, we use geofacet to visualize monthly rainfall patterns and temperature across various weather stations in Singapore. Each panel represents a station's total rainfall or mean temperature, positioned approximately according to its actual location, allowing users to intuitively observe spatial differences and trends.

## 1. Monthly Mean Temperature:

```{r}
# Aggregate to monthly means
weather_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station",
       x = "Month", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Monthly Total Rainfall:

```{r}
# Aggregate to monthly total rainfall
rainfall_monthly <- weather_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Total Monthly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet plot
ggplot(rainfall_monthly, aes(x = Month, y = `Total Monthly Rainfall (mm)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Total Monthly Rainfall by Station",
       x = "Month", y = "Total Monthly Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3. Mean Temperature Trend (2020-2024):

```{r}
# Aggregate to yearly means
weather_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_yearly, aes(x = Year, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +  
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Mean Temperature by Station",
       x = "Year", y = "Mean Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Total Rainfall Trend (2020-2024):

```{r}
# Aggregate to yearly total rainfall
rainfall_yearly <- weather_data %>%
  mutate(Year = year(Date)) %>%
  group_by(Station, Region, Year) %>%
  summarise(`Total Yearly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart of yearly total rainfall
ggplot(rainfall_yearly, aes(x = factor(Year), y = `Total Yearly Rainfall (mm)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Yearly Total Rainfall by Station",
       x = "Year", y = "Total Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Mean Monthly Temperature by Station (2024)

Filter for desired year (eg. 2024).

```{r}
# Filter for 2024 and aggregate to monthly means
weather_2024_monthly <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

# Geofacet plot for 2024
ggplot(weather_2024_monthly, aes(x = Month, y = `Mean Temperature (°C)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Mean Temperature by Station (2024)",
       x = "Month", y = "Mean Temperature (°C)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 6. Monthly Total Rainfall by Station (2024)

Filter for desired year (eg. 2024).

```{r}
# Aggregate to monthly total rainfall for 2024
rainfall_2024 <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Daily Rainfall Total (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart with abbreviated month labels
ggplot(rainfall_2024, aes(x = Month, y = `Daily Rainfall Total (mm)`, color = Region)) +
  geom_line(size = 1) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Total Rainfall by Station (2024)",
       x = "Month", y = "Total Rainfall (mm)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Aggregate to monthly total rainfall for 2024
rainfall_2024 <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Station, Region, Month) %>%
  summarise(`Daily Rainfall Total (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Geofacet bar chart with abbreviated month labels
ggplot(rainfall_2024, aes(x = Month, y = `Daily Rainfall Total (mm)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Total Rainfall by Station (2024)",
       x = "Month", y = "Total Rainfall (mm)") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 7. Monthly Climatology (Average Across Years)

This plot answers the question: "What’s a typical temperature like in each month for each part of Singapore?"

```{r}
weather_climatology <- weather_data %>%
  filter(year(Date) %in% 2020:2024) %>%
  mutate(Month_Label = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Station, Region, Month_Label) %>%
  summarise(`Avg Temp (2020–2024)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

ggplot(weather_climatology, aes(x = Month_Label, y = `Avg Temp (2020–2024)`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Monthly Climatology (2020–2024) by Station",
       x = "Month", y = "Avg Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))
```

## 8. Extreme Rainfall Events Count

Filter for 2024 (or a desired range), then define an extreme rainfall threshold (e.g., \> 80mm). This plot shows the frequency of days with heavy rainfall (\>80mm, for example).

```{r}

# Define threshold
extreme_threshold <- 80

# Prepare base data with month labels
extreme_rainfall <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month"),
         Month_Label = factor(format(Month, "%b"), levels = month.abb)) %>%
  filter(`Daily Rainfall Total (mm)` > extreme_threshold) %>%
  group_by(Station, Region, Month_Label) %>%
  summarise(`Extreme Rainfall Days` = n(), .groups = "drop") %>%
  complete(Station, Region, Month_Label = factor(month.abb, levels = month.abb), fill = list(`Extreme Rainfall Days` = 0))

# Plot
ggplot(extreme_rainfall, aes(x = Month_Label, y = `Extreme Rainfall Days`, fill = Region)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Extreme Rainfall Events by Station (2024)",
       subtitle = paste0("Days with rainfall > ", extreme_threshold, "mm"),
       x = "Month", y = "Number of Extreme Rainfall Days") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))

```

## 9. Dry vs Wet Days Per Month

```{r}
# Categorize days as 'Dry' or 'Wet' and count per month
dry_wet_days <- weather_data %>%
  filter(year(Date) == 2024) %>%
  mutate(Month = floor_date(Date, "month"),
         Month_Label = factor(format(Month, "%b"), levels = month.abb),
         RainType = ifelse(`Daily Rainfall Total (mm)` > 0, "Wet", "Dry")) %>%
  group_by(Station, Region, Month_Label, RainType) %>%
  summarise(Days = n(), .groups = "drop")

# Geofacet stacked bar chart
ggplot(dry_wet_days, aes(x = Month_Label, y = Days, fill = RainType)) +
  geom_col() +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_manual(values = c("Dry" = "skyblue", "Wet" = "steelblue")) +
  theme_minimal() +
  labs(title = "Dry vs Wet Days Per Month by Station (2024)",
       x = "Month", y = "Number of Days", fill = "Day Type") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))

```

## 10. Rainfall Anomalies (Z-score) plot

```{r}
# Step 1: Aggregate to monthly total rainfall
rainfall_monthly <- weather_data %>%
  mutate(Year = year(Date),
         Month = floor_date(Date, "month"),
         Month_Label = month(Month, label = TRUE, abbr = TRUE)) %>%
  group_by(Station, Region, Month, Year, Month_Label) %>%
  summarise(`Monthly Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

# Step 2: Calculate monthly mean & SD for each Station + Month_Label (climatology)
rainfall_stats <- rainfall_monthly %>%
  group_by(Station, Month_Label) %>%
  summarise(
    Mean_Rainfall = mean(`Monthly Rainfall (mm)`, na.rm = TRUE),
    SD_Rainfall = sd(`Monthly Rainfall (mm)`, na.rm = TRUE),
    .groups = "drop"
  )

# Step 3: Join and compute Z-score
rainfall_anomalies <- rainfall_monthly %>%
  left_join(rainfall_stats, by = c("Station", "Month_Label")) %>%
  mutate(Z_Score = (`Monthly Rainfall (mm)` - Mean_Rainfall) / SD_Rainfall)

# Step 4: Plot Z-scores for a specific year (e.g., 2024)
rainfall_anomalies_2024 <- rainfall_anomalies %>%
  filter(Year == 2024)

# Plot
ggplot(rainfall_anomalies_2024, aes(x = Month_Label, y = Z_Score, fill = Z_Score > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"), labels = c("Drier", "Wetter")) +
  facet_geo(~ Station, grid = custom_grid) +
  theme_minimal() +
  labs(title = "Rainfall Anomalies by Station (2024)",
       subtitle = "Z-score of Monthly Rainfall vs Historical Mean",
       x = "Month", y = "Rainfall Z-score", fill = "Anomaly") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))
```

## 11. Hextile Map (Total Rainfall for 2024)

```{r}
rainfall_total_2024 <- weather_data %>%
  filter(lubridate::year(Date) == 2024) %>%
  group_by(Station, Region) %>%
  summarise(`Total Rainfall (mm)` = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE), .groups = "drop")

ggplot(rainfall_total_2024, aes(x = 1, y = 1, fill = `Total Rainfall (mm)`)) +
  geom_tile(color = "white") +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_void() +
  labs(title = "Total Rainfall by Station (2024)",
       fill = "Rainfall (mm)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        legend.position = "bottom")
```

## 12. Hextile Map (Mean Temperature for 2024)

```{r}
temperature_mean_2024 <- weather_data %>%
  filter(lubridate::year(Date) == 2024) %>%
  group_by(Station, Region) %>%
  summarise(`Mean Temperature (°C)` = mean(`Mean Temperature (°C)`, na.rm = TRUE), .groups = "drop")

library(ggplot2)
library(geofacet)

ggplot(temperature_mean_2024, aes(x = 1, y = 1, fill = `Mean Temperature (°C)`)) +
  geom_tile(color = "white") +
  facet_geo(~ Station, grid = custom_grid) +
  scale_fill_gradient(low = "mistyrose", high = "darkred") +
  theme_void() +
  labs(title = "Mean Temperature by Station (2024)",
       fill = "Temperature (°C)") +
  theme(strip.text = element_text(size = 9, face = "bold"),
        legend.position = "bottom")
```

# Spatial Interpolation

Spatial interpolation is a technique used to estimate unknown values at specific geographic locations based on known values from surrounding areas. In this module, we apply spatial interpolation to predict rainfall at locations where no weather stations are present, using data from nearby stations. This allows us to generate a continuous surface of estimated rainfall across Singapore, filling in spatial gaps and providing a more comprehensive view of rainfall distribution. By integrating spatial interpolation into our visual analytics application, users can gain deeper insights into localized weather patterns and better understand the spatial dynamics of rainfall.

## Importing Shapefile

```{r}
mpsz2019 <- st_read(dsn = "data/geospatial", 
                    layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

## User-defined variables

In this prototype, we compute the total monthly rainfall from Daily Rainfall Total (mm) field for Feb 2024. On the shiny app, users will be able to select the variable and the month and year.

```{r}
# Summarise rainfall
rain_summary <- weather_data %>%
  mutate(Date = ymd(Date)) %>%
  filter(month(Date) == 2 & year(Date) == 2024) %>%
  group_by(Station) %>%
  summarise(MONTHSUM = sum(`Daily Rainfall Total (mm)`, na.rm = TRUE)) %>%
  ungroup()

# Extract station coordinates
station_coords <- weather_data %>%
  select(Station, Latitude, Longitude) %>%
  distinct()

# Join rainfall summary with coordinates
rfdata <- left_join(rain_summary, station_coords, by = "Station")


```

```{r}
glimpse(rfdata)

```

## Convert to Geospatial Data

```{r}
rfdata_sf <- st_as_sf(rfdata, coords = c("Longitude", "Latitude"), crs = 4326)
```

## Visualising the data

```{r}
tmap_mode("view")
tm_shape(rfdata_sf) +
  tm_dots(fill = "red")
```

```{r}
tmap_mode("view")
tm_shape(mpsz2019) +
  tm_borders() +
tm_shape(rfdata_sf) +
  tm_dots(fill = 'MONTHSUM')
```

## Inverse Distance Weighted (IDW)

```{r}
rfdata_sf <- st_transform(rfdata_sf, crs = 3414)
```

```{r}

# Create a grid raster over the study area
grid <- terra::rast(mpsz2019, nrows = 690, ncols = 1075)

# Generate coordinates for each grid cell
xy <- terra::xyFromCell(grid, 1:ncell(grid))

# Convert coordinates to sf points
coop <- st_as_sf(as.data.frame(xy), coords = c("x", "y"), crs = 3414)

# Clip grid to area inside mpsz2019
coop <- st_filter(coop, mpsz2019)

```

```{r}
idw_model <- gstat(formula = MONTHSUM ~ 1,
                   locations = rfdata_sf,
                   nmax = 5,           
                   set = list(idp = 2)) # Inverse distance power

# Predict rainfall at grid points
coop <- st_transform(coop, crs = st_crs(rfdata_sf))  
idw_result <- predict(idw_model, coop)

```

```{r}
idw_result$x <- st_coordinates(idw_result)[, 1]
idw_result$y <- st_coordinates(idw_result)[, 2]
idw_result$pred <- idw_result$var1.pred

# Rasterize predicted values back to the grid
idw_raster <- terra::rasterize(idw_result, grid, field = "pred", fun = "mean")

```

```{r}

tmap_mode("plot")

tm_shape(idw_raster) +
  tm_raster(
    col = "mean",  
    col.scale = tm_scale_continuous(values = "brewer.blues"),
    col.alpha = 0.7,
    col.legend = tm_legend(title = "Rainfall (mm)")
  ) +
  tm_shape(mpsz2019) +
  tm_borders() +
  tm_title("IDW Interpolation of Feb 2024 Rainfall") +
  tm_layout(legend.outside = TRUE)

```

## Kriging

```{r}

# Create a gstat object
g <- gstat(formula = MONTHSUM ~ 1, data = rfdata_sf)

# Compute experimental variogram
v_emp <- variogram(g)

# Fit a theoretical variogram model 
v_model <- fit.variogram(v_emp, model = vgm("Sph"))  

```

```{r}
# Create Kriging model using fitted variogram
krig_model <- gstat(formula = MONTHSUM ~ 1, 
                    data = rfdata_sf, 
                    model = v_model)

# Predict on grid points (coop)
krig_result <- predict(krig_model, coop)

```

```{r}
krig_result$x <- st_coordinates(krig_result)[, 1]
krig_result$y <- st_coordinates(krig_result)[, 2]
krig_result$pred <- krig_result$var1.pred

# Rasterize to match the `grid` object
krig_raster <- terra::rasterize(krig_result, grid, field = "pred", fun = "mean")
names(krig_raster) <- "kriged_rainfall"

```

```{r}
tmap_mode("plot")

tm_shape(krig_raster) +
  tm_raster(
    col = "kriged_rainfall",
    col.scale = tm_scale_continuous(values = "brewer.blues"),
    col.legend = tm_legend(title = "Rainfall (mm)")
  ) +
  tm_shape(mpsz2019) +
  tm_borders() +
  tm_title("Kriging Interpolation of Feb 2024 Rainfall") +
  tm_layout(legend.outside = TRUE)

```

# References

<https://ryanhafen.com/blog/geofacet/>

<https://r4va.netlify.app/chap24>
