---
title: "Take-home Exercise 1: Exploring Ship Performance: Financial Insights via Visualisation"
author: "Kenneth Yong"
date: "February 1, 2025"
date-modified: "last-modified"
format: 
  html: 
    code-fold: true
    code-summary: Show the code
---

# Overview

## Background

The maritime industry is a key driver of global trade, directly impacting supply chains and economic growth. Efficient ship performance management is essential for maximizing profitability, reducing operational costs, and maintaining competitive advantages. By analyzing revenue generation, cost structures, and profitability across different ship types and routes, businesses can optimize fleet management strategies. Additionally, understanding operational efficiency and seasonal impacts allows companies to make data-driven decisions, improve resource allocation, and enhance overall financial performance, ultimately driving sustainable growth and profitability.

## Objectives

As a graphical editor of a media company, this article aims to:

1.  **Optimize Financial Performance**: Analyze revenue, cost structures, and profitability across ship types and routes to identify key drivers of financial success.
2.  **Enhance Operational Efficiency**: Examine operational metrics to discover cost-saving opportunities and improve fleet utilization.
3.  **Inform Strategic Decision-Making**: Leverage data insights to guide investment decisions, route planning, and capacity management for sustainable growth.

# Getting Started

## Installing and loading packages

The following R packages will be used:

-   **tidyverse:** A collection of R packages designed for data manipulation, exploration, and visualization, providing a consistent and easy-to-use syntax.

-   **plotly**: R package for creating interactive and dynamic data visualizations.

-   **ggrepel:** Enhances ggplot2 plots by preventing text labels from overlapping, ensuring clear and readable annotations.

-   **ggthemes:** Offers additional themes and color schemes for ggplot2, enabling more visually appealing and customized plots.

-   **patchwork:** Facilitates combining multiple ggplot2 plots into a single layout, making it easy to create complex multi-panel visualizations.

-   **ggridges:** Creates ridge plots (or joy plots) to visualize the distribution of a continuous variable across multiple categories.

-   **ggdist:** Provides tools for visualizing distributions and uncertainty, including interval plots and density plots.

-   **scales:** Allows customization of axis labels, breaks, and scales in ggplot2, improving plot readability and presentation.

```{r}
pacman::p_load(tidyverse, plotly,
               ggrepel, ggthemes,
               ggridges, ggdist,
               patchwork, scales)
```

# The Dataset

## About this dataset

The Ship Performance Dataset is a synthetic, yet realistic collection of data designed to represent key operational metrics and attributes of various ship types operating in the Gulf of Guinea. It includes a wide range of variables such as speed, engine power, maintenance status, cargo weight, operational costs, revenue, and efficiency. This dataset can be found on Kaggle at this [link](https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset/data).

**Key Information:**

-   File Format: CSV

-   Number of Rows: 2736

-   Number of Columns: 18

**Column Information:**

Categorical Variables:

-   Ship_Type

-   Route_Type

-   Engine_Type

-   Maintenance_Status

-   Weather_Condition

Continuous Variables:

-   Date

-   Speed_Over_Ground_knots

-   Engine_Power_kW

-   Distance_Traveled_nm

-   Draft_meters

-   Cargo_Weight_tons

-   Operational_Cost_USD

-   Revenue_per_Voyage_USD

-   Turnaround_Time_hours

-   Efficiency_nm_per_kWh

-   Seasonal_Impact_Score

-   Weekly_Voyage_Count

-   Average_Load_Percentage

## Importing the data

Dataset is initialised to `df`.

```{r}
df <- read_csv("data/Ship_Performance_Dataset.csv")
```

## Data Preparation

A brief look at the data using `glimpse().`

```{r}
glimpse(df)
```

### Checking for duplicates

By using `duplicated`, we can confirm that there are no duplicated entries in the dataset.

```{r}
df[duplicated(df),]
```

### Checking for missing values

By running the code chunk below, we can confirm that there are no missing values in each column.

```{r}
colSums(is.na(df))
```

### Data Manipulation

This code chunk below uses the `mutate()` function to create a new column called Profit in the data frame `df`. It calculates profit by subtracting Operational_Cost_USD from Revenue_per_Voyage_USD for each row. This operation preserves the existing columns while adding the new Profit column.

```{r}
df <- df %>%
  mutate(Profit = Revenue_per_Voyage_USD - Operational_Cost_USD)
```

The following code chunks prepare monthly aggregated data for each ship type by converting `Date` into a monthly format, grouping the data by month, and calculating average values for `Revenue_per_Voyage_USD`, `Operational_Cost_USD`, and `Profit` while ignoring missing values. It then converts the month back to a date format for consistent plotting, enabling trend analysis of revenue, cost, and profit over time.

**Overall**

```{r}
df_monthly <- df %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE),
            Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))
```

**Bulk Carrier**

```{r}
df_bulk_monthly <- df %>%
  filter(Ship_Type == "Bulk Carrier") %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE),
            Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))
```

**Container Ship**

```{r}
df_container_monthly <- df %>%
  filter(Ship_Type == "Container Ship") %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE),
            Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))
```

**Fish Carrier**

```{r}
df_fish_monthly <- df %>%
  filter(Ship_Type == "Fish Carrier") %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE),
            Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))
```

**Tanker**

```{r}
df_tanker_monthly <- df %>%
  filter(Ship_Type == "Tanker") %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE),
            Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))
```

Final glimpse at the datasets:

```{r}
glimpse(df)
```

```{r}
glimpse(df_monthly)
```

```{r}
glimpse(df_bulk_monthly)
```

```{r}
glimpse(df_container_monthly)
```

```{r}
glimpse(df_fish_monthly)
```

```{r}
glimpse(df_tanker_monthly)
```

# EDA 1: Distribution of profit across ship types and shipping routes

```{r}
# Exclude Ship_Type and Route_Type = "None"
df_filtered1 <- df %>% filter(Ship_Type != "None")
df_filtered2 <- df %>% filter(Route_Type != "None")

# Plot
p1 <- ggplot(df_filtered1, aes(x = Ship_Type, y = Profit, fill = Ship_Type)) +
  geom_boxplot() +
  labs(title = "Distribution of Profit Across Ship Types", 
       x = "Ship Type", 
       y = "Profit (USD)") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(df_filtered2, aes(x = Route_Type, y = Profit, fill = Route_Type)) +
  geom_boxplot() +
  labs(title = "Distribution of Profit Across Route Types", 
       x = "Route Type", 
       y = "Profit (USD)") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p1, tooltip = "y")

ggplotly(p2, tooltip = "y")
```

::: {.callout-note icon="false"}
## Insights

-   The median profit (middle line of each box) appears similar across all ship types and shipping routes, with Fish Carriers and Long Haul being marginally more profitable than the rest.

-   All ship and route types exhibit a wide interquartile range (IQR), indicating significant variability in profit. This suggests that while some voyages are highly profitable, others operate at much lower margins.

-   The lower whiskers extend below zero for all ship types, meaning that some voyages result in financial losses. This is due to high operational costs and low revenue.

-   The upper whiskers suggest that some voyages achieve much higher-than-average profits. These might correspond to optimized routes, high cargo loads, or cost-efficient operations.

-   Since the profit distributions and spread are quite similar, no single ship and route type appear to have a significantly better or worse risk-reward profile compared to the others.
:::

# EDA 2: Revenue, Operational Cost and Profit Trends Over Time

**An overview of the overall monthly profit trend:**

```{r}
df_monthly <- df %>%
  mutate(Month = format(Date, "%Y-%m")) %>%
  group_by(Month) %>%
  summarize(Average_Profit = mean(Profit, na.rm = TRUE)) %>%
  mutate(Month = as.Date(paste0(Month, "-01")))

pp <- ggplot(df_monthly, aes(x = Month, y = Average_Profit, group = 1, 
                            text = paste("Month:", format(Month, "%b %Y"), 
                                         "<br>Avg Profit:", round(Average_Profit, 2)))) +
  geom_line(color = "steelblue") +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Monthly Trend of Profit",
       x = "Month",
       y = "Average Profit (USD)") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(pp, tooltip = "text")
```

**Revenue, Operational Cost and Profit trend lines stacked:**

```{r}
# Aggregate by month 

df_monthly <- df %>% mutate(Month = format(Date, "%Y-%m")) %>% group_by(Month) %>% summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE), Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE), Average_Profit = mean(Profit, na.rm = TRUE)) %>% mutate(Month = as.Date(paste0(Month, "-01")))

# Revenue Plot

pr1 <- ggplot(df_monthly, aes(x = Month, y = Average_Revenue)) + geom_line(color = "green") + labs(title = "Monthly Revenue Trend", x = "Month", y = "Revenue (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Operational Cost Plot

po1 <- ggplot(df_monthly, aes(x = Month, y = Average_Cost)) + geom_line(color = "red") + labs(title = "Monthly Operational Cost Trend", x = "Month", y = "Operational Cost (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Profit Plot

pp1 <- ggplot(df_monthly, aes(x = Month, y = Average_Profit)) + geom_line(color = "steelblue") + labs(title = "Monthly Profit Trend", x = "Month", y = "Profit (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

pr1 / po1 / pp1
```

::: {.callout-note icon="false"}
## Insights

-   **October 2023 and January to February 2024** show the lowest profit levels while March 2024 shows the highest profit level.

-   The dip in **October 2023** is more attributed to high operational cost while the dips in **January to February 2024** are attributed to low revenue.

-   Conversely, the peak profit in **March 2024** is attributed to high revenue and extremely low operational cost (lowest month in the data)
:::

# EDA 3: Revenue, Operational Cost and Profit trend lines Across Different Ships

::: panel-tabset
## Bulk Carrier

```{r}
# Revenue Trend
p1 <- ggplot(df_bulk_monthly, aes(x = Month, y = Average_Revenue)) + geom_line(color = "green") + labs(title = "Revenue Trend for Bulk Carrier", x = "Month", y = "Revenue (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Operational Cost Trend
p2 <- ggplot(df_bulk_monthly, aes(x = Month, y = Average_Cost)) + geom_line(color = "red") + labs(title = "Operational Cost Trend for Bulk Carrier", x = "Month", y = "Operational Cost (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Profit Trend
p3 <- ggplot(df_bulk_monthly, aes(x = Month, y = Average_Profit)) + geom_line(color = "steelblue") + labs(title = "Profit Trend for Bulk Carrier", x = "Month", y = "Profit (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 / p3
```

## Container Ship

```{r}
# Revenue Trend
p1 <- ggplot(df_container_monthly, aes(x = Month, y = Average_Revenue)) + geom_line(color = "green") + labs(title = "Revenue Trend for Container Ship", x = "Month", y = "Revenue (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Operational Cost Trend
p2 <- ggplot(df_container_monthly, aes(x = Month, y = Average_Cost)) + geom_line(color = "red") + labs(title = "Operational Cost Trend for Container Ship", x = "Month", y = "Operational Cost (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Profit Trend
p3 <- ggplot(df_container_monthly, aes(x = Month, y = Average_Profit)) + geom_line(color = "steelblue") + labs(title = "Profit Trend for Container Ship", x = "Month", y = "Profit (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 / p3
```

## Fish Carrier

```{r}
# Revenue Trend
p1 <- ggplot(df_fish_monthly, aes(x = Month, y = Average_Revenue)) + geom_line(color = "green") + labs(title = "Revenue Trend for Fish Carrier", x = "Month", y = "Revenue (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Operational Cost Trend
p2 <- ggplot(df_fish_monthly, aes(x = Month, y = Average_Cost)) + geom_line(color = "red") + labs(title = "Operational Cost Trend for Fish Carrier", x = "Month", y = "Operational Cost (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Profit Trend
p3 <- ggplot(df_fish_monthly, aes(x = Month, y = Average_Profit)) + geom_line(color = "steelblue") + labs(title = "Profit Trend for Fish Carrier", x = "Month", y = "Profit (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 / p3


```

## Tanker

```{r}
# Revenue Trend
p1 <- ggplot(df_tanker_monthly, aes(x = Month, y = Average_Revenue)) + geom_line(color = "green") + labs(title = "Revenue Trend for Tanker", x = "Month", y = "Revenue (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Operational Cost Trend
p2 <- ggplot(df_tanker_monthly, aes(x = Month, y = Average_Cost)) + geom_line(color = "red") + labs(title = "Operational Cost Trend for Tanker", x = "Month", y = "Operational Cost (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Profit Trend
p3 <- ggplot(df_tanker_monthly, aes(x = Month, y = Average_Profit)) + geom_line(color = "steelblue") + labs(title = "Profit Trend for Tanker", x = "Month", y = "Profit (USD)") + scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 / p3


```
:::

::: {.callout-note icon="false"}
## Insights

-   The ship types have their different seasonal trends.

-   For Bulk Carrier, the most profitable months are August and May.

-   For Container Ships, the most profitable months are November and January.

-   For Fish Carriers, the most profitable month is September with dips in August and October.

-   For Tankers, the most profitable months are the middle of the year.
:::

# EDA 4: Average Revenue versus Operational Cost by Ship Type

```{r}
# Calculate the average Revenue and Operational Cost for each Ship Type
df_avg <- df %>%
  filter(Ship_Type %in% c("Bulk Carrier", "Container Ship", "Fish Carrier", "Tanker")) %>%
  group_by(Ship_Type) %>%
  summarize(Average_Revenue = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
            Average_Cost = mean(Operational_Cost_USD, na.rm = TRUE))

# Scatterplot of Average Revenue vs Average Operational Cost
p <- ggplot(df_avg, aes(x = Average_Cost, y = Average_Revenue, color = Ship_Type)) +
  geom_point(size = 4) +
  labs(title = "Average Revenue vs Average Operational Cost by Ship Type", 
       x = "Average Operational Cost (USD)", 
       y = "Average Revenue per Voyage (USD)") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

ggplotly(p)
```

::: {.callout-note icon="false"}
## Insights

-   **Fish Carriers** generate the highest revenue with moderate operational costs, making them the most cost-efficient.

-   **Container Ships** have high revenue but also the highest operational costs, potentially impacting profitability.

-   **Bulk Carriers** maintain a balanced revenue-to-cost ratio, indicating stable financial performance.

-   **Tankers** have the lowest revenue and operational costs, suggesting lower profit margins.
:::

# EDA 5: Average Profit by Ship Type and Engine Type

```{r}
# Calculate average profit by Engine_Type and Ship_Type
df_avg_profit <- df %>%
  filter(Ship_Type %in% c("Bulk Carrier", "Container Ship", "Fish Carrier", "Tanker")) %>%
  group_by(Engine_Type, Ship_Type) %>%
  summarize(Average_Profit = mean(Profit, na.rm = TRUE), .groups = "drop")

df_avg_profit <- df_avg_profit %>%
  filter(Engine_Type != "None")

# Set custom colors for Engine_Type
engine_colors <- c(
  "Diesel" = "#FF9999",
  "Heavy Fuel Oil (HFO)" = "#99CCFF",
  "Steam Turbine" = "#99FF99")

# Create the bar chart
p <- ggplot(df_avg_profit, aes(x = Ship_Type, y = Average_Profit, fill = Engine_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Profit by Ship Type and Engine Type", 
       x = "Ship Type", 
       y = "Average Profit (USD)") +
  scale_fill_manual(values = engine_colors) +
  scale_y_continuous(labels = comma) +
  theme_minimal()

ggplotly(p)
```

::: {.callout-note icon="false"}
## Insights

-   For Bulk Carriers, Diesel has the highest profitability with Steam Turbine the lowest.

-   For Container Ships, Steam Turbine has the highest profitability with Diesel the lowest.

-   For Fish Carriers, Diesel has the highest profitability with HFO the lowest.

-   For Tankers, it is relatively even across the board but Steam Turbine edges out Diesel marginally with HFO the lowest.
:::

# EDA 6: Average efficiency by Ship Type and Engine Type

```{r}
# Filter out "None" for Ship_Type and Engine_Type
df_filtered <- df %>%
  filter(Ship_Type != "None", Engine_Type != "None")

# Calculate average efficiency by Ship_Type and Engine_Type
df_avg_efficiency <- df_filtered %>%
  group_by(Ship_Type, Engine_Type) %>%
  summarize(Average_Efficiency = mean(Efficiency_nm_per_kWh, na.rm = TRUE), .groups = "drop")

# Create the bar chart
p <- ggplot(df_avg_efficiency, aes(x = Ship_Type, y = Average_Efficiency, fill = Engine_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Efficiency by Ship Type and Engine Type", 
       x = "Ship Type", 
       y = "Average Efficiency (nm/kWh)") +
  theme_minimal()

# Display the plot
ggplotly(p)
```

::: {.callout-note icon="false"}
## Insights

-   Efficiency is relatively even across all ship types and engine types.
:::

# Conclusion

This study analyzes ship profitability, operational costs, and revenue across ship types, routes, and seasonal trends. **Fish Carriers** are the most cost-efficient, while **Container Ships** face profitability challenges due to high operational costs. **Bulk Carriers** maintain a stable revenue-to-cost balance, and **Tankers** operate with thinner profit margins.

**Seasonal trends significantly impact profitability**, with **low profits in October 2023 and January–February 2024** due to high costs and low revenue, while **March 2024 saw peak profits** driven by high revenue and low costs. Each ship type has distinct profitability peaks, emphasizing the need for **strategic fleet and route planning**.

**Diesel engines are the most profitable for Bulk Carriers and Fish Carriers, while Steam Turbines lead for Container Ships.** Tankers show minimal differences across engine types. **Efficiency remains consistent across ships and engines**, suggesting that engine selection depends on affordability and use case.

No single ship type, route, or engine type has a clear financial advantage, highlighting the need for **cost optimization and strategic revenue management**. Future research should examine **fuel price impact, cargo load variations, and route-specific performance metrics** to enhance decision-making.
